---
title: "Codebook"
author: "Rezwan Ahmed"
date: "Friday, June 19, 2015"
output:
  html_document:
    keep_md: yes
---

## The Data
The experiments have been carried out with a group of 30 volunteers within an 
age bracket of 19-48 years. Each person performed six activities (WALKING, 
WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a 
smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer 
and gyroscope, we captured 3-axial linear acceleration and 3-axial angular 
velocity at a constant rate of 50Hz. The experiments have been video-recorded 
to label the data manually. The obtained dataset has been randomly partitioned 
into two sets, where 70% of the volunteers was selected for generating the 
training data and 30% the test data. 

Data for the project: 
https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip 

Original description of the dataset: 
http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

## The variables 

The following list of features are captured as part of the data to evaluate
each subjects activity.

```{r}
features <- read.table("UCI HAR Dataset/features.txt")
features[2]
```


The following list of activities were performed:
```{r}
activities <- read.table("UCI HAR Dataset/activity_labels.txt", header = FALSE)
activities[2]
```


## Transformations to clean up the data 

The following are the steps taken to clean up the data:

```{r,echo=FALSE}
library(data.table)
library(dplyr)
```

#### 1. Merge the training and the test sets to create one data set.

Reading all train data
```{r}
trainSubject <- read.table("UCI HAR Dataset/train/subject_train.txt", header = FALSE)
trainX <- read.table("UCI HAR Dataset/train/X_train.txt", header = FALSE)
trainY <- read.table("UCI HAR Dataset/train/y_train.txt", header = FALSE)
```

Reading all test data
```{r}
testSubject <- read.table("UCI HAR Dataset/test/subject_test.txt", header = FALSE)
testX <- read.table("UCI HAR Dataset/test/X_test.txt", header = FALSE)
testY <- read.table("UCI HAR Dataset/test/y_test.txt", header = FALSE)
```

Merging train and test data for each item
```{r}
allSubject <- rbind(trainSubject, testSubject)
allX <- rbind(trainX, testX)
allY <- rbind(trainY, testY)
```

Loading the feature (X) names and activity (Y) labels and using the second 
column of the features data frame as the feature labels
```{r}
features <- read.table("UCI HAR Dataset/features.txt")
labelX <- t(features[2])
```

Setting all the column labels
```{r}
colnames(allX) <- labelX
colnames(allY) <- "ActivityID"
colnames(allSubject) <- "SubjectID"
```

The combined dataset
```{r}
combinedData <- cbind(allSubject,allX,allY)
```

The dimenstion of the combined data:
```{r}
dim(combinedData)
```

#### 2. Extract only the measurements on the mean and standard deviation for each measurement. 

First separate the features with mean and std. dev in their name
```{r}
cnames <- names(allX)
extCols <- grep(".*Mean.*|.*Std.*", cnames, , ignore.case=TRUE)
extX <- allX[,extCols]
```

Subset of the data
```{r}
subsetData <- cbind(allSubject, extX, allY)
```

The dimenstion of the extracted data:
```{r}
dim(subsetData)
```

#### 3. Use descriptive activity names to name the activities in the data set 

Loading the activity labels included in the dataset
```{r}
activities <- read.table("UCI HAR Dataset/activity_labels.txt", header = FALSE)
nactivity <- nrow(activities)
activities[2]
```

Replacing the Activity IDs from the dataset with actual activity labels.
```{r}
tempActIDs <- as.character(subsetData$ActivityID)
for (i in 1:nactivity){
    tempActIDs[tempActIDs[] == i] <- as.character(activities[i,2])
}
subsetData$ActivityID <- as.factor(tempActIDs)
```


Changing the name of the column from ActivityID to Activity
```{r}
ncols <- ncol(subsetData)
colnames(subsetData)[ncols] <- "Activity"
```

The different types of activities captured as a factor of the Activity column
```{r}
levels(subsetData$Activity)
```

#### 4. Appropriately labels the data set with descriptive variable names.  

Original labels of the features:
```{r}
names(subsetData)
```

Replacing "t" with "Time"
```{r}
names(subsetData) <- gsub("^t", "Time", names(subsetData))
```

Replacing "f" with "Freq"
```{r}
names(subsetData) <- gsub("^f", "Freq", names(subsetData))
```

Replacing "Mag" with "Magnitude"
```{r}
names(subsetData) <- gsub("Mag", "Magnitude", names(subsetData))
```

Replacing "BodyBody" with "Body"
```{r}
names(subsetData) <- gsub("BodyBody", "Body", names(subsetData))
```

Replacing "Acc" with "Accelerometer"
```{r}
names(subsetData) <- gsub("Acc", "Accelerometer", names(subsetData))
```

Replacing "tBody" with "TimeBody"
```{r}
names(subsetData) <- gsub("tBody", "TimeBody", names(subsetData))
```

Updated labels of the features:
```{r}
names(subsetData)
```

#### 5. From the data set in step 4, create a second, independent tidy data set with the average of each variable for each activity and each subject. 

```{r}
tidyData<-aggregate(. ~SubjectID + Activity, subsetData, mean)
tidyData<-tidyData[order(tidyData$SubjectID,tidyData$Activity),]
head(tidyData, n=10)
```

Saving the tidy data to a file.
```{r}
write.table(tidyData, file="TidyData.txt", row.names = FALSE)
```